---
title: "1. Annotating texts with LLM"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This section demonstrates the core functionality of `tidychat`: annotating text with multiple draws and quantifying uncertainty. We'll analyze therapy session notes and identify which cases need human review.

## Setup

First, load the package and create a chat constructor. We'll use Ollama with a small free model:

```{r, warning = F, message = F}
library(tidychat)
library(tidyverse)

# Initialize Ollama (make sure Ollama is running locally)
# Run: ollama pull llama3.2:3b
chat <- choose_ollama(model = "llama3", temperature = 0.3, seed = 123)
```

## Prepare Data

We have five therapy session notes to analyze:

```{r, warning = F, message = F}
therapy_notes <- data.frame(
  client_id = 1:5,
  session_note = c(
    "Client reported significant progress this week. Has been using coping 
     strategies consistently and feeling more in control of anxiety symptoms.",
    
    "Client expressed frustration with lack of progress. Feels stuck and 
     unmotivated despite practicing techniques.",
    
    "Mixed session. Some good moments discussing family dynamics, but client 
     seemed distracted and fatigued throughout.",
    
    "Breakthrough session! Client connected childhood experiences to current 
     relationship patterns for the first time. Emotional but hopeful.",
    
    "Routine check-in. Client is stable but notes feeling 'blah' about everything. 
     No significant changes positive or negative."
  ),
  stringsAsFactors = FALSE
)
```

## Annotate with Multiple Draws

We'll classify each note as positive, negative, or neutral, taking 5 independent draws per client. Multiple draws allow us to measure how confident the LLM is in each classification.

The key arguments to `llm_annotate()` are:

- **`col`**: The name of the new column that will contain the annotations (`sentiment` in this example)
- **`input`**: The column containing the text to be analyzed (`session_note`)
- **`id`**: An optional unique identifier column; if not provided, a `.id` column is automatically created
- **`prompt_task`**: The instruction given to the LLM describing what to extract
- **`extract_type`**: The type of output expected – here "category" tells the LLM to choose from predefined options
- **`categories`**: The allowed values for categorical extraction, which the LLM is constrained to return
- **`chat`**: The chat constructor function that defines which model and parameters to use
- **`n_draws`**: The number of independent draws (repeated judgments) per item – the core feature for uncertainty quantification

```{r, eval = F}
results <- therapy_notes %>%
  llm_annotate(
    col = sentiment,
    input = session_note,
    id = client_id,
    prompt_task = "Classify the sentiment of this therapy note as positive, 
                    negative, or neutral. Return only one word.",
    extract_type = "category",
    categories = c("positive", "negative", "neutral"),
    chat = chat,
    n_draws = 5,
    show_progress = TRUE
  )

head(results)
```

```{r, echo = F, warning = F, message = F}
results <- read_csv('res1.csv')
head(results)
```

The output of `llm_annotate()` is a tidy data frame where each row represents a single draw from the LLM. The key columns include: `item_id` (the original identifier), `draw_id` (the draw number, from 1 to `.n_draws`), the annotation column (named after `col`, in this case `sentiment`), plus metadata columns `model_name`, `provider`, `prompt`, `params_hash`, and individual parameter columns (`temperature`, `max_tokens`, `seed`). This long-format structure is designed to work with `dplyr` functions such as `group_by()` and others. 

## Quantify Uncertainty (or Agreement)

We can then use `summarise_draws()` to quantify uncertainty across the 5 draws. By specifying `sentiment` as the column to summarize, the function collapses the multiple draws per item into a single row of summary statistics, grouped by the identifiers we provided (`id` and `session_note`).

```{r}
uncertainty <- results %>%
  group_by(id, session_note) %>%
  summarise_draws(sentiment)

print(uncertainty)
```

The output of `summarise_draws()` is a tidy summary table where each row corresponds to one original item (identified by the `id` and `session_note` columns), collapsing the 5 draws per item into summary statistics. The majority column shows the most frequent `sentiment` across draws (all 5 draws agreed in this case), `agreement` reports the proportion of draws matching that majority (1.0 = perfect agreement), `entropy` quantifies uncertainty (0 = no uncertainty), and `distribution` shows the complete frequency distribution across all draws (e.g., "positive (5)" indicates all 5 draws were positive).

## Flag Cases for Human Review

Here, we establish so that cases with agreement below 70% (the default threshold) should be reviewed by a human.

```{r}
needs_review <- uncertainty %>%
  filter(uncertain == TRUE) %>%
  select(id, majority, agreement, entropy, distribution)

print(needs_review)
```

As expected, it is empty because all LLM responses have agreement of 1. 

## Summary

`tidychat`'s multi-draw approach provides a principled way to quantify LLM uncertainty. By taking multiple independent judgments and measuring agreement, users can make informed decisions about which automated annotations to trust.
