---
title: "3. Choosing LLM Providers and Configuring Models"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This section shows how to extract structured data from research abstracts using `llm_extract()`. This is particularly useful for systematic reviews and meta-analyses where researchers need to quickly code hundreds of abstracts.

## Introduction

Before you can annotate text or extract structured data, you need to create a chat constructor that tells `tidychat` which LLM provider and model to use, along with key parameters like `temperature`, `top_p`, and `max_tokens`. This section explains the available providers, how to configure them, and how to choose the right settings for your use.

## Why Multiple Providers?

Different research tasks have different requirements:

```{r}
knitr::kable(
  data.frame(
    Provider = c("Ollama", "Groq", "Gemini", "OpenAI", "Anthropic", "DeepSeek", "Azure OpenAI"),
    Best_For = c("Local, private, free", "Fast inference, free tier", "Multimodal, free tier",
                 "State-of-the-art quality", "Safety, long context", "Cost-effective, reasoning",
                 "Enterprise, compliance"),
    Cost = c("Free", "Free / Paid", "Free / Paid", "Paid", "Paid", "Paid (free credits)", "Paid"),
    API_Key_Required = c("No", "Yes", "Yes", "Yes", "Yes", "Yes", "Yes")
  ),
  caption = "Overview of LLM providers supported by tidychat",
  align = c("l", "l", "l", "c")
)
```


## Local Models with Ollama (No API Key Required)

Ollama runs LLMs locally on your machine, making it ideal for prototyping, private data, and situations where you cannot send data to cloud APIs.

### Installation

First, install Ollama from ollama.ai, then pull a model in bash:

```{bash, eval = F}
ollama pull llama3
```

### Basic Usage

```{r, eval = F}
library(tidychat)

# Create a chat constructor with default settings
chat <- choose_ollama(model = "llama3")

# With custom parameters
chat <- choose_ollama(
  model = "llama3.2:3b",
  temperature = 0.3,
  max_tokens = 500
)
```

## Key Parameters for All Providers

```{r}
knitr::kable(
  data.frame(
    Parameter = c("temperature", "top_p", "max_tokens", "seed"),
    Description = c(
      "Controls randomness. Lower = more deterministic, higher = more creative",
      "Nucleus sampling. Considers only the most probable tokens whose cumulative probability exceeds p",
      "Maximum length of generated response",
      "Random seed for reproducibility (Ollama only)"
    ),
    Typical_Range = c("0-1", "0-1", "50-4096", "integer"),
    When_to_Use = c(
      "Use 0 for reproducible coding tasks; 0.5-1 for exploring variability",
      "Lower (0.5) = more focused; higher (0.95) = more diverse",
      "Set higher for long-form extraction, lower for classification",
      "Set to a fixed value for deterministic results"
    )
  ),
  caption = "Key parameters for LLM configuration in tidychat",
  align = c("l", "l", "c", "l")
)
```

### Understanding Temperature

Temperature is the single most important parameter for research applications:

```{r, eval = F}
# Deterministic - perfect for reproducible coding but biased
chat_deterministic <- choose_ollama(model = "llama3", temperature = 0)

# Balanced - default for most tasks
chat_balanced <- choose_ollama(model = "llama3", temperature = 0.3)

# Creative - useful for exploration or generating variability
chat_creative <- choose_ollama(model = "llama3", temperature = 0.8)
```

Practical guidance:
- temperature = 0: Every draw produces the same result. Use for deterministic coding tasks.
- temperature = 0.3-0.5: Moderate variability. Default for most research.
- temperature = 0.7-1.0: High variability. Use when you want to measure uncertainty across draws.

## Cloud Providers (API Keys Required)

### Google Gemini

Gemini offers a generous free tier and strong multimodal capabilities, including native PDF support.

```{r, eval = F}
# Use Flash model for speed (free tier)
chat <- choose_gemini(model = "gemini-2.5-flash", api_key =  "your-api-key")

# Use Pro for complex tasks
chat <- choose_gemini(model = "gemini-2.5-pro", temperature = 0.1, api_key =  "your-api-key")
```

Get an API key: Visit Google AI Studio → Get API key → Free tier available.

### OpenAI

OpenAI's GPT models are state-of-the-art but require payment.

```{r, eval = F}
# Use GPT-4o Mini for cost-effective annotation
chat <- choose_openai(model = "gpt-4o-mini", temperature = 0.3, api_key =  "your-api-key")

# Use GPT-4o for complex extraction
chat <- choose_openai(model = "gpt-4o", max_tokens = 4096)
```

Get an API key: Visit OpenAI Platform → Create API key → Requires billing.

### Anthropic Claude

Claude models excel at safety and long-context understanding.

```{r, eval = F}
# Use Haiku for fast, cheap annotation
chat <- choose_anthropic(model = "claude-3-haiku-20240307", api_key =  "your-api-key")

# Use Sonnet for balanced performance
chat <- choose_anthropic(model = "claude-3-5-sonnet-20241022", api_key =  "your-api-key")
```

Get an API key: Visit Anthropic Console → Create API key → Requires billing.

### Groq

Groq offers a generous free tier.

```{r, eval = F}
# Use Llama 3 70B for high-quality extraction
chat <- choose_groq(model = "llama3-70b-8192", temperature = 0, api_key =  "your-api-key")

# Use Mixtral for speed/cost balance
chat <- choose_groq(model = "mixtral-8x7b-32768", api_key =  "your-api-key")
```

Get an API key: Visit Groq Console → Create API key → Free tier available.

### DeepSeek

DeepSeek offers competitive pricing and strong performance.

```{r, eval = F}
chat <- choose_deepseek(model = "deepseek-chat")

# Or via OpenRouter (alternative access)
Sys.setenv(OPENROUTER_API_KEY = "your-openrouter-key", api_key =  "your-api-key")
chat <- choose_deepseek(
  model = "deepseek/deepseek-chat",
  base_url = "https://openrouter.ai/api/v1",
  api_key =  "your-api-key"
)
```
